\documentclass[sigconf]{acmart}

\usepackage{booktabs} % For formal tables
\usepackage{cleveref}

\usepackage{xspace}
\usepackage{listings}
\usepackage{listingconfig}
\usepackage{array}
\usepackage{balance}

\lstset{language=Racket}

\newcommand{\etal}{\mbox{et al.}\xspace}
\newcommand{\htdp}{\mbox {\sc htdp}\xspace}

\renewenvironment{quote}%
 {\list{}{\leftmargin=0.25in\rightmargin=0.25in}\item[]}%
 {\endlist}

\newcommand{\sthree}{\mbox{\sc StudA}\xspace}
%\newcommand{\sfour}{\mbox{\sc StudB}\xspace}
\newcommand{\ssix}{\mbox{\sc StudB}\xspace}
\newcommand{\sseven}{\mbox{\sc StudC}\xspace}
\newcommand{\sten}{\mbox{\sc StudD}\xspace}
%\newcommand{\stwelve}{\mbox{\sc StudD}\xspace}

\newcommand{\rainsummary}[2]{{\bf Correctness}: #1\par \noindent {\bf Overall Structure}: #2\vskip5pt}
\newcommand{\inlinequote}[1]{\emph{#1}}

\newtheorem{obs}{Observation}
\newtheorem{openq}{Open Question}

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}


% DOI
\acmDOI{10.475/123_4}

% ISBN
\acmISBN{123-4567-24-567/08/06}

%Conference
\copyrightyear{2017}
\acmYear{2017}
\setcopyright{acmlicensed}
\acmConference{ICER'17}{}{August 18--20, 2017, Tacoma, WA, USA.}
\acmPrice{15.00}
\acmDOI{http://dx.doi.org/10.1145/3105726.3106183}
\acmISBN{ISBN 978-1-4503-4968-0/17/08}

\begin{document}
\title{Sometimes, Rainfall Accumulates: \\ Talk-Alouds with
  Novice Functional Programmers}

\author{Kathi Fisler}
%\orcid{1234-5678-9012}
\affiliation{%
  \institution{Brown University}
  \streetaddress{Department of Computer Science}
  \city{Providence}
  \state{RI}
  \postcode{02912-1910}
}
\email{kfisler@cs.brown.edu}

\author{Francisco Enrique Vicente Castro}
%\orcid{1234-5678-9012}
\affiliation{%
  \institution{WPI}
  \streetaddress{Department of Computer Science}
  \city{Worcester}
  \state{MA}
  \postcode{01609}
}
\email{fgcastro@cs.wpi.edu}

% The default list of authors is too long for headers}
\renewcommand{\shortauthors}{Fisler and Castro}


\begin{abstract}
  When functional programming is used in studies of the Rainfall
  problem in CS1, most students seem to perform fairly well.  A
  handful of students, however, still struggle, though with different
  surface-level errors than those reported for students programming
  imperatively.  Prior research suggests that novice programmers
  tackle problems by refining a high-level program schema that they
  have seen for a similar problem.  Functional-programming students,
  however, have often seen multiple schemas that would apply to
  Rainfall. How do novices navigate these choices?  This paper
  presents results from a talk-aloud study in which novice functional
  programmers worked on Rainfall.  We describe the criteria that drove
  students to select, and sometimes switch, their high-level program
  schema, as well as points where students realized that their chosen
  schema was not working.  Our main contribution lies in our
  observations of how novice programmers approach a multi-task planning
  problem in the face of multiple viable schemas.
\end{abstract}

\copyrightyear{2017} 
\acmYear{2017} 
\setcopyright{acmlicensed}
\acmConference{ICER '17}{August 18-20, 2017}{Tacoma, WA, USA}\acmPrice{15.00}\acmDOI{10.1145/3105726.3106183}
\acmISBN{978-1-4503-4968-0/17/08}
\fancyhead{}
\settopmatter{printacmref=false, printfolios=false}
%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.
%
% \begin{CCSXML}
% <ccs2012>
% <concept>
% <concept_id>10003456.10003457.10003527.10003531.10003533</concept_id>
% <concept_desc>Social and professional topics~Computer science education</concept_desc>
% <concept_significance>500</concept_significance>
% </concept>
% <concept>
% <concept_id>10003456.10003457.10003527.10003531.10003533.10011595</concept_id>
% <concept_desc>Social and professional topics~CS1</concept_desc>
% <concept_significance>500</concept_significance>
% </concept>
% <concept>
% <concept_id>10003456.10003457.10003527.10003540</concept_id>
% <concept_desc>Social and professional topics~Student assessment</concept_desc>
% <concept_significance>500</concept_significance>
% </concept>
% <concept>
% <concept_id>10003120.10003121.10003122.10003334</concept_id>
% <concept_desc>Human-centered computing~User studies</concept_desc>
% <concept_significance>300</concept_significance>
% </concept>
% </ccs2012>
% \end{CCSXML}

%\ccsdesc[500]{Social and professional topics~Computer science education}
%\ccsdesc[500]{Social and professional topics~CS1}
%\ccsdesc[500]{Social and professional topics~Student assessment}
%\ccsdesc[300]{Human-centered computing~User studies}

\keywords{Rainfall; program schemas; functional programming}

\maketitle

\input{icer-body}

\bibliographystyle{ACM-Reference-Format}
\balance
\bibliography{rainfall}

\end{document}
